---
title: "HW6 Joining Tables"
author: "Gerard Duncan"
format: html
editor: visual
---

---
title: "Homework 6 - Joining Tables and Software Tips"
format: docx
editor: visual
---

Fill out [this Quarto document](https://github.com/BIFX552-22/CourseInfo/hw/06joining_tables.qmd) while completing the lesson on graphics, vectorization and functions in R.

```{r}
#| include: false
if(FALSE)
{
    # use this code to install the package we're using for this lesson
    BiocManager::install("UCLouvain-CBIO/rWSBIM1207")
}

library("rWSBIM1207")
data(jdf)
```

### [Joining Tables](https://carpentries-incubator.github.io/bioc-intro/50-joining-tables/index.html)

::: callout-tip
## Key Points

-   Join tables of data in R
:::

#### Challenge 1

Using the `full_join` function, join tables `jdf4` and `jdf5`. What has happened for observations `P26039` and `P02468`?

> ```{r}
> # Insert code here

full_join(jd4, jd5)
> ```

#### Challenge 2

Join tables `jdf4` and `jdf5`, keeping only observations in `jdf4`.

```{r}
# Insert code here

left_join(jdf5)
```

#### Challenge 3

Join tables `jdf4` and `jdf5`, keeping only observations in `jdf5`.

```{r}
# Insert code here
right_join(jdf5)
```

#### Challenge 4

Join tables `jdf4` and `jdf5`, keeping observations observed in both tables.
```{r}
inner_join(jdf4, jdf5)
```


#### Challenge 5

Given the inner join,

```{r}
inner_join(jdf6, jdf7)
```

interpret the result, where both tables have duplicated keys.

\<insert answer here\>